Imperative Programming with Variational Effects

### 2018-07-23 paper/paper.tex
more work

\usepackage{setspace} 
\singlespacing
\usepackage{graphicx}                                        
\usepackage{amssymb}                                         
\usepackage{amsmath}                                         
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithmic}
\usepackage{url}
\usepackage{tikz}         
\usetikzlibrary{matrix}
\usepackage{listings}
\usepackage{tabu}
\usepackage{mathptmx}
\usepackage{fullpage}
\usepackage{enumerate}
\usepackage{fancyvrb}
\usepackage{mathpartir}

@@

%\def \one {\inferrule
%  {one \\ two \\ three \\ or \\ more \\ premises}
%  {and \\ any \\ number \\ of \\ conclusions \\ as \\ well}
%}
%$$
%\fbox {\vbox {\advance \hsize by -2\fboxsep \advance \hsize by -2\fboxrule
%       \linewidth\hsize
%        $$\one $$}}
%$$

@@

\newcommand{\GetCtx}[2]{\mathit{ctx}(#1,#2)}
\def \T {\mathit{true}}
\def \F {\mathit{false}}
\def \BigVErr {\infer [VS-Err] { } {(C,S,E,\T,s) \Downarrow_{VS} (S,E,\T)}}
\def \BigVSkip {\infer [VS-Skip] { } {(C,S,E,F,\text{skip}) \Downarrow_{VS} (S,E,F)}}
\def \BigVAssn {\infer [VS-Assn] {(\GetCtx{C}{E},S,e) \Downarrow_{VA} v} {(C,S,E,\F,x \coloneqq e) \Downarrow_{VS} (S[(\GetCtx{C}{E},x) \rightarrow v],E,\F)}}
\def \BigVSeq {\infer [VS-Seq] {(C,S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C,S',E',F,s') \Downarrow_{VS} (S'',E'',F')} {(C,S,E,s\ \text{seq}\ s') \Downarrow_{VS} (S'',E'',F')}}
\def \BigVIfOne {\infer [VS-If1] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'}) \\ (C \wedge \mathit{whenFalse}(v),S',E',\F) \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S'',E'',F \wedge F')}}
\def \BigVIfTwo {\infer [VS-If2] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E}) \\ (C \wedge \mathit{whenFalse}(v),S,E,\F) \Downarrow_{VS} (S',E',F)} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S',E',F)}}
\def \BigVIfThree {\infer [VS-If3] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'})} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S',E',F)}}
\def \BigVIfFour {\infer [VS-If4] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'})} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S,E,\F)}}
\def \BigVWhileT {\infer [VS-WhileT] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C \wedge \mathit{whenTrue}(v),S',E',F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S'',E'',F')}}
\def \BigVWhileF {\infer [VS-WhileF] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E})} {(C,S,E,\F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S,E,\F)}}
\def \BigVThrow {\infer [VS-Throw] {(\GetCtx{C}{E},S,e) \Downarrow_{VA} v} {(C,S,E,\F,\text{throw}\ e) \Downarrow_{VS} (S,\mathit{combine}((C,v),E),\T)}}

\begin{mathpar}
\BigVErr \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIfOne \and
\BigVIfTwo \and
\BigVIfThree \and
\BigVIfFour \and
\BigVWhileT \and
\BigVWhileF \and
\BigVThrow
\end{mathpar}

@@

All resources are maintained in a global resource environment during execution of a program
in the Resource DSL. This environment is a mapping from paths to resource values. Paths are
analogous to the Unix filesystem paths that most programmers are already familiar with, allowing for
simple organization of resources in treelike structures. For example, a system may place its server
bandwidth resource at \prog{/Server/Bandwidth}, while placing its client bandwidth at \prog{/Client/Bandwitdth}.

@@

Statements are used in order to sequence and combine the above effects into programs in the Resource DSL.
One particular difficulty arises in determining how to represent a variational block of statements. Typically we
would use a simple list of statements, but in the variational setting we must ask what kind of variational list
we should use. Fortunately, recent work \cite{karltodo} has determined \emph{segment lists} to be a particularly
efficient implementation of variational linked lists. A segment list is a list of segments, where segments are either
plain lists or a choice between segment lists. Therefore, we define our blocks to be segment lists of
statements.

@@

data Stmt
     = Do Path Effect       -- ^ apply an effect
     | If (V Expr) Block Block  -- ^ conditional statement
     | In Path Block        -- ^ do work in a sub-environment
     | For Var (V Expr) Block   -- ^ loop over indexed sub-environments
     | Let Var (V Expr) Block   -- ^ extend the variable environment
     | Load (V Expr) [V Expr] -- ^ load a sub-model or profile
\end{program}

>>>>

\usepackage{setspace} 
\singlespacing
\usepackage{graphicx}                                        
\usepackage{amssymb}                                         
\usepackage{amsmath}                                         
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithmic}
\usepackage{url}
\usepackage{tikz}         
\usetikzlibrary{matrix}
\usepackage{listings}
\usepackage{tabu}
\usepackage{mathptmx}
\usepackage{fullpage}
\usepackage{enumerate}
\usepackage{fancyvrb}
\usepackage{mathpartir}
\usepackage{syntax}

@@

@@

\newcommand{\GetCtx}[2]{\mathit{ctx}(#1,#2)}
\def \T {\mathit{true}}
\def \F {\mathit{false}}
\def \BigVErr {\infer [VS-Err] { } {(C,S,E,\T,s) \Downarrow_{VS} (S,E,\T)}}
\def \BigVSkip {\infer [VS-Skip] { } {(C,S,E,F,\text{skip}) \Downarrow_{VS} (S,E,F)}}
\def \BigVAssn {\infer [VS-Assn] {(\GetCtx{C}{E},S,e) \Downarrow_{VA} v} {(C,S,E,\F,x \coloneqq e) \Downarrow_{VS} (S[(\GetCtx{C}{E},x) \rightarrow v],E,\F)}}
\def \BigVSeq {\infer [VS-Seq] {(C,S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C,S',E',F,s') \Downarrow_{VS} (S'',E'',F')} {(C,S,E,s\ \text{seq}\ s') \Downarrow_{VS} (S'',E'',F')}}
\def \BigVIfOne {\infer [VS-If1] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'}) \\ (C \wedge \mathit{whenFalse}(v),S',E',\F) \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S'',E'',F \wedge F')}}
\def \BigVIfTwo {\infer [VS-If2] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E}) \\ (C \wedge \mathit{whenFalse}(v),S,E,\F) \Downarrow_{VS} (S',E',F)} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S',E',F)}}
\def \BigVIfThree {\infer [VS-If3] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'})} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S',E',F)}}
\def \BigVIfFour {\infer [VS-If4] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'})} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S,E,\F)}}
\def \BigVWhileT {\infer [VS-WhileT] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C \wedge \mathit{whenTrue}(v),S',E',F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S'',E'',F')}}
\def \BigVWhileF {\infer [VS-WhileF] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E})} {(C,S,E,\F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S,E,\F)}}
\def \BigVThrow {\infer [VS-Throw] {(\GetCtx{C}{h:t},S,e) \Downarrow_{VA} v} {(C,S,h:t,\F,\text{throw}\ e) \Downarrow_{VS} (S,((C,v):h):t,\T)}}
\def \BigVTry {\infer [VS-Try] {(C,S,[]:E,\F,s) \Downarrow_{VS} (S',[]:E,\F)} {(C,S,E,\F,\text{try}\ s\ \text{catch}\ x\ \text{in}\ s') \Downarrow_{VS} (S',E,\F)}}
\def \BigVCatch {\infer [VS-Catch] {(C,S,[]:E,\F,s) \Downarrow_{VS} (S',(h:t):E,F) \\ (C \wedge \mathit{catchCtx}(h:t),S'[(\GetCtx{C}{E},x) \rightarrow \mathit{combine}(h:t)],E,\F,s']) \Downarrow_{VS} (S'',E',F')} {(C,S,E,\F,\text{try}\ s\ \text{catch}\ x\ \text{in}\ s') \Downarrow_{VS} (S''[(\T,x) \rightarrow S'(\T,x)],E',F')}}
\def \BigVChcOne {\infer [VS-Chc1] {\mathit{sat}(\GetCtx{C}{E} \wedge A) \\ \mathit{sat}(\GetCtx{C}{E} \wedge \neg A) \\ (C \wedge A,S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C \wedge \neg A,S',E',\F,s') \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S'',E'',F \wedge F')}}
\def \BigVChcTwo {\infer [VS-Chc2] {\mathit{unsat}(\GetCtx{C}{E} \wedge A) \\ \mathit{sat}(\GetCtx{C}{E} \wedge \neg A) \\ (C \wedge \neg A,S,E,\F,s') \Downarrow_{VS} (S',E',F)} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S',E',F)}}
\def \BigVChcThree {\infer [VS-Chc3] {\mathit{sat}(\GetCtx{C}{E} \wedge A) \\ \mathit{unsat}(\GetCtx{C}{E} \wedge \neg A) \\ (C \wedge A,S,E,\F,s) \Downarrow_{VS} (S',E',F)} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S',E',F)}}
\def \BigVChcFour {\infer [VS-Chc4] {\mathit{unsat}(\GetCtx{C}{E} \wedge A) \\ \mathit{unsat}(\GetCtx{C}{E} \wedge \neg A)} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S,E,\T)}}

\begin{mathpar}
\BigVErr \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIfOne \and
\BigVIfTwo \and
\BigVIfThree \and
\BigVIfFour \and
\BigVWhileT \and
\BigVWhileF \and
\BigVThrow \and
\BigVTry \and
\BigVCatch
\end{mathpar}
\begin{mathpar}
\BigVChcOne \and
\BigVChcTwo \and
\BigVChcThree \and
\BigVChcFour
\end{mathpar}

@@

All resources are maintained in a global resource environment during execution of a program
in the Resource DSL. This environment is a mapping from paths to resource values. Paths are
analogous to the Unix filesystem paths that most programmers are already familiar with, allowing for
simple organization of resources in treelike structures. For example, a system may place its server
bandwidth resource at \prog{/Server/Bandwidth}, while placing its client bandwidth at \prog{/Client/Bandwidth}.
Again, by analogy to filesystem paths, resource paths can be either absolute or relative. Relative paths are
evaluated with respect to the execution environment's current path. For example, a reference to the resource
at the relative path \prog{../GPU/Speed} would evaluate to \prog{/Servers/Hardware/GPU/Speed}
given an execution environment with \prog{/Servers/Hardware/CPU} as its current path.

@@

Statements are used in order to sequence and combine the above effects into programs in the Resource DSL.
The most basic statement is the \prog{Do} command, which takes a path to a resource and applies an effect to that
path. The other statements are a collection of control structures that allow for the sequencing and construction of
programs out of these \prog{Do} commands. \prog{If} and \prog{For} are straightforward control structures that allow
for conditional execution and looping, respectively. \prog{In} executes a block in the context of a particular path, where all
relative paths in an \prog{In} block are evaluated with reference to the supplied path value. This allows
users to easily describe work that should be done in particular resource sub-environments. \prog{Let} allows for the
extension of the current variable environment at the statement level. Finally, \prog{Load} permits the calling of named
subroutines with a list of arguments.

One particular difficulty arises in determining how to represent a variational block of statements. Typically we
would use a simple list of statements, but in the variational setting we must ask what kind of variational list
we should use. Fortunately, recent work \cite{karltodo} has determined \emph{segment lists} to be a particularly
efficient implementation of variational linked lists. A segment list is a list of segments, where segments are either
plain lists or a choice between segment lists. Therefore, we define our blocks to be segment lists of
statements.

@@

data Stmt
     = Do Path Effect
     | If (V Expr) Block Block
     | In Path Block
     | For Var (V Expr) Block
     | Let Var (V Expr) Block
     | Load (V Expr) [V Expr]
\end{program}
